---
title: "HW 4"
author: "Collin Register"
date: "12/29/2023"
output: 
  html_document:
    number_sections: true
---

This homework is designed to give you practice fitting a logistic regression and working with statistical/philosophical measures of fairness.  We will work with the `titanic` dataset which we have previously seen in class in connection to decision trees.  

Below I will preprocess the data precisely as we did in class.  You can simply refer to `data_train` as your training data and `data_test` as your testing data.  




```{r}

#this is all of the preprocessing done for the decision trees lecture.  

path <- 'https://raw.githubusercontent.com/guru99-edu/R-Programming/master/titanic_data.csv'
titanic <-read.csv(path)
head(titanic)

library(dplyr)

#replace ? with NA
replace_question_mark <- function(x) {
  if (is.character(x)) {
    x <- na_if(x, "?")
  }
  return(x)
}

titanic <- titanic %>%
  mutate_all(replace_question_mark)

set.seed(678)
shuffle_index <- sample(1:nrow(titanic))
head(shuffle_index)

titanic <- titanic[shuffle_index, ]
head(titanic)

library(dplyr)
# Drop variables
clean_titanic <- titanic %>%
select(-c(home.dest, cabin, name, x, ticket)) %>% 
#Convert to factor level
    mutate(pclass = factor(pclass, levels = c(1, 2, 3), labels = c('Upper', 'Middle', 'Lower')),
    survived = factor(survived, levels = c(0, 1), labels = c('No', 'Yes'))) %>%
na.omit()
#previously were characters
clean_titanic$age <- as.numeric(clean_titanic$age)
clean_titanic$fare <- as.numeric(clean_titanic$fare)
glimpse(clean_titanic)

create_train_test <- function(data, size = 0.8, train = TRUE) {
    n_row = nrow(data)
    total_row = size * n_row
    train_sample <- 1: total_row
    if (train == TRUE) {
        return (data[train_sample, ])
    } else {
        return (data[-train_sample, ])
    }
}
data_train <- create_train_test(clean_titanic, 0.8, train = TRUE)
data_test <- create_train_test(clean_titanic, 0.8, train = FALSE)

```

#
Create a table reporting the proportion of people in the training set surviving the Titanic.  Do the same for the testing set.  Comment on whether the current training-testing partition looks suitable.  

```{r}
round(prop.table(table(data_train$survived)),2)
data_train$survived<-as.factor(data_train$survived)

round(prop.table(table(data_test$survived)),2)
data_test$survived<-as.factor(data_test$survived)



```

*The current training-testing partition does look suitable because they seem to be representative of each other.*

#
Use the `glm` command to build a logistic regression on the training partition.  `survived` should be your response variable and `pclass`, `sex`, `age`, `sibsp`, and `parch` should be your response variables.  

```{r}

model<-glm(survived~pclass+sex+age+sibsp+parch,family=binomial(link = 'logit'),data=data_train)
summary(model)

```

We would now like to test whether this classifier is *fair* across the sex subgroups.  It was reported that women and children were prioritized on the life-boats and as a result survived the incident at a much higher rate.  Let us see if our model is able to capture this fact.  

#

Subset your test data into a male group and a female group.  Then, use the `predict` function on the male testing group to come up with predicted probabilities of surviving the Titanic for each male in the testing set.  Do the same for the female testing group.  

```{r}

#student input
male <- subset(data_test, sex == "male")
female <- subset(data_test, sex == "female")


fitted.results.male<-predict(model,newdata=male,type='response')
fitted.results.male
mean(fitted.results.male)

fitted.results.female<-predict(model,newdata=female,type='response')
fitted.results.female
mean(fitted.results.female)

```

# 

Now recall that for this logistic *regression* to be a true classifier, we need to pair it with a decision boundary.  Use an `if-else` statement to translate any predicted probability in the male group greater than $0.5$ into `Yes` (as in Yes this individual is predicted to have survived).  Likewise an predicted probability less than $0.5$ should be translated into a `No`.  

Do this for the female testing group as well, and then create a confusion matrix for each of the male and female test set predictions.  You can use the `confusionMatrix` command as seen in class to expidite this process as well as provide you necessary metrics for the following questions.  

```{r}
library(caret)
fitted.results.male.Y.N<-ifelse(fitted.results.male>0.5,"Yes","No")
fitted.results.female.Y.N<-ifelse(fitted.results.female>0.5,"Yes","No")

matrix_male <- confusionMatrix(as.factor(fitted.results.male.Y.N), male$survived,positive = "Yes")
matrix_male

matrix_female <- confusionMatrix(as.factor(fitted.results.female.Y.N), female$survived,positive = "Yes")
matrix_female

```

#
We can see that indeed, at least within the testing groups, women did seem to survive at a higher proportion than men (24.8\% to 76.3\% in the testing set).  Print a summary of your trained model and interpret one of the fitted coefficients in light of the above disparity.  

```{r}
summary(model)
```

*When looking at the summary of the fitted model, we can observe that the sexmale coefficient matches our observation from the discrepancy in the male and female survival rates above. More specifically, a one unit increase in sexmale results in a -2.684206 unit decrease in the log odds units, holding all else equal.* 


#

Now let's see if our model is *fair* across this explanatory variable.  Calculate five measures (as defined in class) in this question: the Overall accuracy rate ratio between females and males, the disparate impact between females and males, the statistical parity between females and males, and the predictive equality as well as equal opportunity between females and males (collectively these last two comprise equalized odds).  Set a reasonable $\epsilon$ each time and then comment on which (if any) of these five criteria are met.  


```{r}

accuracy_ratio_female_to_male<-((4+59)/(4+59+15+2))/((93+4)/(93+28+4+4))
epsilon=0.2
abs(1-accuracy_ratio_female_to_male)>epsilon
#this one is not met

disparate_impact_female_to_male<-((2+59)/(15+4+2+59))/((28+4)/(28+4+4+93))
epsilon=0.2
disparate_impact_female_to_male<(1-epsilon)
#this one is not met

statistical_parity<-abs(((2+59)/(15+4+2+59))-((28+4)/(28+4+4+93)))
epsilon=0.2
statistical_parity>epsilon
#this one is met

#combine to make equliazed odds
predictive_equality<-abs((2/(4+2))-(28/(28+93)))
epsilon=0.2
predictive_equality>epsilon
#this one is not met

equal_opportunity<-abs((59/(15+59))-(4/(4+4)))
epsilon=0.2
equal_opportunity>epsilon
#this one is met
```

*For the accuracy ratio, the criteria is not met because the ratio deviates from 1 by more than epsilon. For the disparate impact the criteria is not met because the impact is not less than 1 minus epsilon. For the statistical parity measure,the criteria is met because it is greater than epsilon. For the predictive equality measure the criteria is not met but for the equal opportunity measure it is. Combining these two to make equalized odds,the criteria is not met because both conditions must be satisfied*.  

It is always important for us to interpret our results in light of the original data and the context of the analysis.  In this case, it is relevant that we are analyzing a historical event post-facto and any disparities across demographics identified are unlikely to be replicated.  So even though our model fails numerous of the statistical fairness criteria, I would argue we need not worry that our model could be misused to perpetuate discrimination in the future.  After all, this model is likely not being used to prescribe a preferred method of treatment in the future.  


#

Even so, provide a *philosophical* notion of justice or fairness that may have motivated the Titanic survivors to act as they did. Spell out what this philosophical notion or principle entails?

*A notion of fairness that may have motivated the Titanic survivors to act as they did may be the difference principle. The idea comes from John Rawls and explains justice as fairness. Under this principle, when differences exist, resources are allocated to protect the most vulnerable.In the case of the Titanic, women were the most vulnerable class and therefore were given more resources(lifeboats, life vests,etc) in order to maintain some level of fairness. The vulnerability here could come from stereotypes of that time period or gender norms.* 


